---
title: 'Data 607 Wk 7: HTML, XML, and JSON'
author: "Aaron Grzasko"
date: "October 9, 2016"
output: 
    html_document:
        theme: simplex
        highlight: haddock
---

---

## Assignment Overview  
  
For this assignment, I first collected five attributes for three different statistics texts on my bookshelf.  
  
Here is a list of those attributes:  

* Author(s)  
* Title  
* Publisher
* Original publication year  
* ISBN 10 number  
  
Next, I manually stored the book attribute data in three separate file formats:  
  
* HTML
* XML
* JSON  
  
In this R Markdown file, I use multiple R packages--most notably *jsonlite* and *XML*--to read each raw file format into R.  
I then store each data file as an R data frame and perform various data cleansing techniques.  The ultimate goal of this exercise is to transform the three separate, raw files into identical data frames.  

## Load Libraries
```{r, comment=NA, warning=FALSE, message=FALSE}
if (!require(XML)) install.packages('XML')
if (!require(stringr)) install.packages('stringr')
if (!require(RCurl)) install.packages('RCurl')
if (!require(knitr)) install.packages('knitr')
if (!require(jsonlite)) install.packages('jsonlite')
```

## HTML Table
  
**Read in data**  
  
Let's first read the HTML table data, which is stored on my Github repository:  

```{r, comment=NA, tidy=TRUE}
# url for raw html file; using getURL() from RCurl package 
html_raw <- getURL("https://raw.githubusercontent.com/spitakiss/Data607_Wk7/master/books.html")  

# read in data, store as data frame; using readHTMLTable() function from the XML package
html_df <- readHTMLTable(html_raw, header=TRUE, as.data.frame = TRUE,trim = TRUE, stringsAsFactors=FALSE) 
html_df <- data.frame(html_df)
```
  
  
**First Look**  
  
Let's examine the initial data frame, as well as various column attributes using the `summary()` function:  
  
```{r, comment=NA}
kable(html_df)

summary(html_df)
```
  
You may have noticed that there are separate columns for *author1* and *author2*.  This is because the raw html file had separate tag elements with these names.  The *author2* column is only populated when a book as more than one author.  
  
There are a few issues with the initial data frame:  
  
* Each column name starts with the string "NULL."  
* We have white space or blank values in the *author2* column.  We wish to populate thee elements with `NA`.  
* The publication year was converted to a character.  I would prefer to store the year as an integer.  
  
**Data Scrubbing**   
  
Below, I perform the necessary data cleanup:

```{r}
# clean up column names 
names(html_df) <- str_replace_all(names(html_df),"NULL.","")

# replace white spaces or blanks with NAs
html_df$author2 <- ifelse(html_df$author2 %in% c(""," ","NA"), NA, html_df$author2)

# convert year to integer
html_df$orig_pub_year <- as.integer(html_df$orig_pub_year)

```
  
**Final Data Frame**  
  
Here is our final data frame created using the original HTML file:  

```{r}
kable(html_df)
```



## XML  
  
**Read in data**  
  
Let's retrieve and read in the xml raw file, stored on my Github repository:  
  
```{r}
# url
xml_raw <- getURL("https://raw.githubusercontent.com/spitakiss/Data607_Wk7/master/books.xml")

# reading to data frame using function from XML package
xml_df <- xmlToDataFrame(xml_raw, stringsAsFactors=FALSE)

```
  
**First Look**  
  
Let's examine the data frame in its current form:  
  
```{r, comment=NA}
kable(xml_df)

summary(xml_df)
```
  
We again have multiple issues with our current data frame:  
  
* There is only one *authors* column instead of two.  This is because the raw XML file had one <authors> tag with two separate attributes, *author1* and *author2*.  This problem presumably could have been avoided if the raw XML file had separate *<author1>* and *<author2>* tags.  
* The publication year was converted to a character.  This should be stored as an integer.  
  

**Data Scrubbing**  
  
I will now address the issues noted in the previous section:  
  
  
```{r}
# use regex from stringr package to  create separate vector objects for author1 and author2 
author1 <- str_sub(xml_df$authors,start = str_locate(xml_df$authors,'author1=[[:print:]]+?\"'))
author1 <- str_replace_all(author1,'"',"")
author1 <- str_replace_all(author1, "author1=","")

author2 <- str_sub(xml_df$authors,start = str_locate(xml_df$authors,'author2=([[:print:]])+"?'))
author2 <- str_replace_all(author2,'"',"")
author2 <- str_replace_all(author2, "author2=","")

# add author 1 and author2 objects to existing data frame
xml_df$author1 <- author1
xml_df$author2 <- author2

# subset data frame, exclude the original authors column
xml_df <- subset(xml_df, select =c(author1, author2, title, publisher, orig_pub_year, isbn10))

# convert publication year data type to integer
xml_df$orig_pub_year <- as.integer(xml_df$orig_pub_year)

```
  
**Final Data Frame**  
  
Here is the final data frame created from the original XML file:  
  
```{r}
kable(xml_df)
```

## JSON  
  
**Read in data**  
  
Let's retrieve and read in the json file, once again stored on my Github repository:  
```{r, comment=NA}
# url
json_raw <- getURL("https://raw.githubusercontent.com/spitakiss/Data607_Wk7/master/books.json")

# store as data frame, using fromJSON() function in jsonlite package
json_df <- fromJSON(json_raw)
```
  
**First Look**  
  
```{r, comment=NA}
json_df

summary(json_df)
```
  
You'll notice that I did not use the `kable()` function with the data frame.  This is because the program errors out when I attempt to do so.   
  
Here are the issues with the data frame:  
  
* The original JSON raw file had a key, *authors*, which itself had two nested keys, *author1* and *author1*.  When we converted the data to an R data frame, the original *authors* column was treated as a nested data frame within the main json_df data frame.  This issue is also prevented us from using the `kable()` function.  
*  The publication year is again stored as a string.  This problem may have been avoided if the raw JSON file did not list each publication year in double quotes.  
  
**Data Scrubbing**  
  
Let's fix issues noted above:  

```{r}

# create new data frame columns, author1 and author2

json_df$author1 <- json_df$authors[[1]]
json_df$author2 <- json_df$authors[[2]]

# subset the data frame to exclude the authors column
json_df <- subset(json_df, select  = c(author1, author2, title, publisher, orig_pub_year, isbn10))


# convert publication year to an integer
json_df$orig_pub_year <- as.integer(json_df$orig_pub_year)


```
  
**Final Data Frame**  
  
Now we have a clean data frame, and we can properly use the `kable()` function:  
  
```{r}
kable(json_df)
```



